Likelihoods for a general class of ARGs under the SMC

### 2025-04-18 review-responses.tex

\begin{minipage}[b]{2.5in}
  Resubmission Cover Letter \\
  {\it Genetics}
\end{minipage}
\hfill
\begin{minipage}[b]{2.5in}
    Gertjan Bisschop  \\
  \today
\end{minipage}

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are writing to submit a revised version of our manuscript,
XXX
titled 
``XXX''.
We appreciate the thoughtful comments by the reviewers.
We've done our best to address all the comments,
as detailed in our Response to Reviewers,
XXX

In addition to a Response to Reviewers (in which page/line numbers refer to the revised manuscript file),
we also provide a pdf with differences to the previously submitted version highlighted.

\vskip 2em

\noindent \hspace{4em}
\begin{minipage}{3in}
\noindent
{\bf Sincerely,}

\vskip 2em

{\bf 
   Gertjan Bisschop, Peter Ralph, and Jerome Kelleher
}\\
\end{minipage}

\vskip 4em

\pagebreak

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
The manuscript by Bisschop et al. presents a dense yet concise derivation of a novel approach for computing the likelihood of a broad class of Ancestral Recombination Graphs (ARGs) under the Sequentially Markovian Coalescent (SMC) model. This is a significant advance that enables the use of output from heuristic ARG inference tools such as tsinfer, ARG-Needle, and Threads, for rigorous statistical inference of population genetic parameters. These recent ARG inference methods can infer local gene trees across genomes for hundreds to thousands of samples, but they do not estimate recombination events directly in their data models, which prevented the application of previous formulations of the likelihood.
\end{quote}

\begin{quote}
The new likelihood framework has the potential to unlock parameter inference-such as recombination rates and historical population sizes-from large-scale genomic datasets using highly efficient methods. The manuscript provides a compelling proof-of-concept for such applications, while appropriately leaving more extensive exploration for future work.
\end{quote}

\begin{quote}
The subject matter is conceptually challenging due to the inherent complexity of the topic, but the authors provide rigorous and accessible explanations. Figures 1 and 2 are especially helpful.
\end{quote}

\begin{point}{}
A Python implementation of the method is available on GitHub. However, due to the algorithm's complexity and the very limited documentation of the available code, it would be beneficial to include pseudocode in the appendix. This pseudocode should lay out the full sequence of steps required to compute the likelihood, ideally supplemented with additional explanatory figures. Doing so would not only enhance the clarity and completeness of the manuscript, but also facilitate implementation in other programming languages.
\end{point}

\reply{
}

\begin{point}{\revref} %line 212
 Since you talk about the rate of coalescence, I think a ``lambda'' is missing in front of the sum (i.e., multiplying by 1/2/N).
\end{point}

\reply{
}

\begin{point}{\revref} % line 218
 Could you provide more detail or intuition for this formula of the likelihood? E.g., what's the cumulative hazard in the context of the coalescent or the ARG.
\end{point}

\reply{
}

\begin{point}{\revref} % line 262
 Again, I think there is a ``lambda'' missing in front of the integral. Alternatively, lambda is meant to be multiplied with the F function, like in equation 4, but then it also needs to be added to the formula in line 265.
\end{point}

\reply{
}

\begin{point}{\revref} % line 577
 I don't understand ``, along with additional information that allows us to sequentially recover (...) efficiently.'' Recover what?
\end{point}

\reply{
}

%%%%%%%%%%%%%%
\reviewersection{2}
\begin{quote}
This paper does a couple of interrelated things.
\begin{enumerate}
    \item The authors develop a new backward-in-time formulation of the Sequentially Markov Coalescent (SMC) model.
    \item This backward-in-time formulation of the SMC model provides a way to efficiently calculate likelihood for a sample of genomes from this model, provided that we have information on how these genomes are related to each other. The latter is done by a data structure called Ancestral Recombination Graph (ARG). The authors provide an algorithm and Python code for this likelihood calculation (is this emphasised enough across the paper!?).
    \item The authors also evaluate the likelihood under different levels of information available in the ARG. This enables them to connect challenges when comparing methods and outputs from the rigorous, yet ``scale-limited'', probabilistic ARG inference methods with scaleable, yet less-rigorous, heuristic ARG inference methods.
\end{enumerate}
I welcome this paper and quite like it's succinctness. I am not able to follow all the maths, but the rationale and presentation is clear to me and I only have minor comments.
\end{quote}

% ABSTRACT

\begin{point}{\revref} % \&75
 Isn't it possible to run SINGER on up to thousands of samples? I know that several hundreds is also a thousand!
\end{point}

\reply{
}

\begin{point}{}
Mention that the standard SMC formulation was left-to-right
\end{point}

\reply{
}

\begin{point}{\revref} % 21:
 for ARG inference that this new formulation opens
\end{point}

\reply{
}

% INTRO

\begin{point}{\revref} % 24:
 evol biol $\to$ genetics?
\end{point}

\reply{
}

\begin{point}{\revref} % 27:
 ancestral processes $\to$ genetic processes?
\end{point}

\reply{
}

\begin{point}{\revref} % 35:
 Bayesian $\to$ probabilistic (I think this is more general and would encompas any model-based approach, say max likelihood etc.; alternatively consider using model-based, but probabilistic/statistical assumes we have an underlying model)
\end{point}

\reply{
}

\begin{point}{\revref} % 47:
 statistical $\to$ probabilistic
\end{point}

\reply{
}

\begin{point}{\revref} % 66:
 and heuristic methods omit these details in different ways. Notablym the sig....
\end{point}

\reply{
}

\begin{point}{\revref} % 72:
 uncertainty of (point) estimates.
\end{point}

\reply{
}

\begin{point}{\revref} % 82:
 heuristic methods
\end{point}

\reply{
}

\begin{point}{\revref} % 84:
 likelihood of a dataset? under a model (to make it more general)
\end{point}

\reply{
}

\begin{point}{\revref}
I would appreciate if you here make it clear which information is missing in these different tools. Sorry for nitpicking.
\end{point}

\reply{
}

\begin{point}{\revref} % 98:
 a new formulation of SMC model and corresponding ARG likelihood
\end{point}

\reply{
}

% RESULTS

\begin{point}{\revref} % 117:
 specific genome (node can also be a sample so best to remove ancestral)
\end{point}

\reply{
}

\begin{point}{\revref} % 118:
 sampled genomes are related at each base pair.
\end{point}

\reply{
}

\begin{point}{\revref} % 121:
 sampled and ancestral haplotypes (haploid genomes), or ...
\end{point}

\reply{
}

\begin{point}{Fig \ref{fig:smc-unary}:}
 it would be cool if you shade regions (in blue) also in the local trees? That's effectively uncertain time around the A, B, and C node!
\end{point}

\reply{
}

\begin{point}{Fig \ref{fig:smc-unary} caption:}
An ARG ``generated under the SMC'' - do you need this last part in my quotes?
\end{point}

\reply{
}

\begin{point}{Figure~\ref{fig:smc-unary} caption:}
with a the remaining $\to$ with the remai....
\end{point}

\reply{
    Thanks! Done.
}

\begin{point}{\revref} % 130:
 either sample nodes or common ancestry nodes
\end{point}

\reply{
}

\begin{point}{\revref} % 138:
 is unary in the second tree.
\end{point}

\reply{
}

\begin{point}{\revref} % 140:
 all lineages with recombination event
\end{point}

\reply{
}

\begin{point}{\revref} % 144:
 sometime between the times of nodes f and h; similarlyt in next line % 145
\end{point}

\reply{
}

\begin{point}{\revref{} and a few lines later} % 156/160:
 consolidate how you write backwards in time
\end{point}

\reply{
}

\begin{point}{\revref{} and elsewhere}
similarly for left-to-right and along-the-genome
\end{point}

\reply{
}

\begin{point}{\revref} % 164:
 had to Google what ``abutting'' means! Consider using bordering/adjoning
\end{point}

\reply{
}

\begin{point}{\revref} % 169:
 I don't fully grasp the meaning of ``tree-valued process'' - tree-generating process or tree-structured process ...
\end{point}

\reply{
}

% Here starts the part that I am not sufficiently-skilled to judge the text, though I could read and follow what is says and means!

\begin{point}{\revref} % 249:
 define ``eligible links'' sooner? Not critical.
\end{point}

\reply{
}

\begin{point}{\revref} % 280:
 ARG likelihood formulation in Eq. (5) ... of nodes (haplotypes that existed at some time t)
\end{point}

\reply{
}

\begin{point}{\revref} % 289:
 with a sample from the SMC model
\end{point}

\reply{
}

\begin{point}{\revref} % 300:
 cite what simplification does (Wang et al., YYYY; Kelleher et al. - your PLOS Genetics paper). Do you need to cite this sooner in the manuscript?
\end{point}

\reply{
}

\begin{point}{\revref} % 300:
 for this application, despite retaininig fewer nodes and edges.
\end{point}

\reply{
}

\begin{point}{\revref{} and subsequent paragraph} % 306-316:
 Is this number of polytomies too small and is then the impact on the likelihood small? It would be good to do one more extreme example for supplement to ensure this. This is the point where you can strengthen your paper the most.
\end{point}

\reply{
}

\begin{point}{Fig \ref{fig:lik-surface} caption:}
 lines overlap hence we cannot see the full red line.
\end{point}

\reply{
}

\begin{point}{\revref{} and elsewhere} % 319:
 Consolidate how you ref to figs (Fig 1. or Fig 3).
\end{point}

\reply{
}

\begin{point}{\revref} % 321:
 you have not mentioned how many mutations the simulation produced. I understand that these are not important for ARG likelihood (emphasise this), but good to tie up things together.
\end{point}

\reply{
}

\begin{point}{\revref} % 328:
 It's not that likelihood function performs well, right!? It's how absence of complete information gives close to the full information likelihood value!
\end{point}

\reply{
}

\begin{point}{\revref} % 332:
 remove significant? It does not look that bad!
\end{point}

\reply{
}

% DISCUSSION

\begin{point}{\revref} % 352:
 modern heuristic ARG ...
\end{point}

\reply{
}

\begin{point}{\revref} % 358:
 consider Kuhner-Yamato-Felsenstein $\to$ KYF (for consistency)
\end{point}

\reply{
}

\begin{point}{\revref} % 381:
 probabilistic approach
\end{point}

\reply{
}

\begin{point}{\revref} % 387:
 probabilistic inference
\end{point}

\reply{
}

\begin{point}{\revref} % 392-394:
 I checked the code and is indeed there! Thanks for sharing! The algorithm has an acronym RUNSMC standing for ``Recombination (time) UNaware SMC''. Should this be emphasised in the manuscript? Hmm, but it's not fully unaware, is it? You do consider topology, so think about a better acronym.
\end{point}

\reply{
}

### 2025-04-18 paper.tex
.

@@

\begin{abstract}
Ancestral recombination graphs (ARGs) are the focus of much ongoing research
interest. Recent progress in inference has made ARG-based approaches feasible
across of range of applications, and many new methods using inferred ARGs as
input have appeared. This progress on the long-standing problem of ARG
inference has proceeded in two distinct directions. First,
the Bayesian inference of ARGs under the Sequentially Markov
Coalescent (SMC), is now practical for tens-to-hundreds of samples. 
Second, approximate models and heuristics can now scale to sample sizes two to three
orders of magnitude larger. Although these heuristic methods are reasonably
accurate under many metrics, one significant drawback is that the ARGs they 
estimate do not have the topological properties required to compute a 
likelihood under models such as the SMC under present-day formulations.
In particular, heuristic inference methods typically do not estimate
precise details about recombination events, which are currently
required to compute a likelihood.
In this paper we present a 
backwards-time formulation of the SMC 
and derive a straightforward definition 
of the likelihood of a general class of ARG under this model. 
We show that this formulation does not require precise details of recombination events
to be estimated, and is robust to the presence of polytomies. 
We discuss the possibilities for inference that this opens.
\end{abstract}

@@

\section*{Introduction}
Following recent breakthroughs in simulation and inference methods,
there is now strong interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary
biology~\citep{lewanski_era_2024,brandt_promise_2024,nielsen_inference_2025}.
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes.
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Numerous applications taking advantage of this detailed history
in inferred ARGs are now appearing~\citep{
stern_approximate_2019,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
ignatieva_ongoing_2022,
wang_complex_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
osmond_estimating_2021,
huang_estimating_2024,
grundler_geographic_2024,
korfmann_simultaneous_2024,
deraje_inferring_2024,
speidel2025high}
and it seems likely that many more will
follow~\citep{harris_database_2019,harris_using_2023}.
Although there are many different approaches to ARG
inference~\citep{wong_general_2023}, two broad classes of method
have been the focus of most recent interest.

% TODO Maybe swap the order of these two paragraphs around, so it's 
% more natrual to talk about the advantages of model-driven stuff next.
The first class of inference methods are Bayesian approaches
that sample ARGs under a population genetics model such as the
coalescent with recombination~\citep{hudson_properties_1983},
and its approximation, the Sequentially Markovian
Coalescent~\citep{mcvean_approximating_2005,marjoram_fast_2006}.
ARGweaver~\citep{rasmussen_genome-wide_2014,hubisz_mapping_2020} 
has been the most widely used and studied,
and until recently consistently out-performed other methods in terms 
of accuracy in a variety of benchmarks~\citep{brandt2022evaluation},
and has successfully been applied in several
contexts~\citep[e.g.,][]{de_chimpanzee_2016,shriner_whole_2018,hejase_genomic_2020,
stankowski_genetic_2024}. 
The recently-introduced method SINGER takes a similar approach,
and by using an improved Monte Carlo algorithm
promises to be even more accurate and substantially more 
scalable~\citep{deng_robust_2024}. 
The second class of inference methods that has been of recent interest 
are based on heuristics and approximate models.
Relate~\citep{speidel_method_2019},
tsinfer~\citep{kelleher_inferring_2019},
ARG-Needle~\citep{zhang_biobank-scale_2023} 
and Threads~\citep{gunnarsson_scalable_2024}
work on quite different 
principles, but share some common properties. 
Firstly, they are all 
heuristic and approximation driven, favouring computational
scalability over a basis in a statistical model.
Secondly, they all produce a single
ARG as output, inferring a deterministic point estimate.
Thirdly, they regard the precise timing of nodes as a separate
problem, focusing only on the relative ordering of nodes when 
producing the initial ARG. General purpose ARG dating methods are 
appearing~\citep{wohns_unified_2022,deng2025general} and can also improve
dating performance in ARGs sampled from the SMC~\citep{deng2025general}.
Fourthly (and most importantly for the 
purposes of this paper) the ARGs that heuristic methods produce 
lack some of the topological information present in the output of ARGweaver or SINGER. 
This occurs, roughly speaking, because of a basic difference in their
approach to uncertainty.

There is considerable uncertainty in ARG inference, and often a fundamental
lack of information in the sequence data to distinguish different
possibilities. Consider, for example, a case in which we have three
ancestral lineages with identical sequences. We know that there
must have been two coalescence events, but there is no mutational information to 
distinguish their relative ordering. Sampling methods overcome this problem
by randomly choosing an ordering, with the uncertainty communicated by the order
changing in the different ARGs sampled from the posterior.
The other approach (used by methods such as tsinfer) is to communicate
the uncertainty structurally by means of a polytomy: we have no
information about the intermediate coalescent event, and so we omit it.
Similarly, there is often a fundamental lack of information about the 
details of recombination events, and heuristic methods (in different ways)
omit these details. In many ways, the significant progress made in
scalability by these heuristic methods is \emph{because} they don't
attempt to precisely reconstruct recombination events.

There are substantial advantages to the statistically rigorous sampling 
approaches of ARGweaver and SINGER.
In particular, they tend to be more accurate (at the 
scale at which they can be compared),
and sampling ARGs from a
posterior distribution provides a means of quantifying the uncertainty 
around estimates.
On the other hand, heuristic methods have the potential
to use the greater amount of information in large datasets,
capturing fine-scale details about the recent past.
% thus possibly producing even more accurate estimates
% (although this has not, to our knowledge, been tested).
Although SINGER is a major step forward
in terms of scalability over ARGweaver and other methods, inference is still 
only feasible for hundreds of samples. 
Meanwhile, tsinfer, ARG-Needle and Threads have 
been applied to datasets of hundreds of \emph{thousands} of samples.
With the explosive growth in dataset
sizes in recent years~\citep{caulfield2017national,turnbull2018100,
bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole,
all2024genomic,stark2024call,cook2025our}
there is a pressing need for more statistical
rigour in large scale ARG inference.
% This inability to connect the results of large-scale inference methods
% with a generative model
% is a fundamental limitation, and limits both the utility of 
% current point estimates as well as the possibilities for further
% inference. Specifically, we are unable to quantify uncertainty 
% around heuristic point estimates. We cannot explore ARG space
% around the point estimates using (e.g., MCMC), as we have no way definition
% of improvement. We cannot compare models of demography, or evaluate the 
% goodness of fit for model estimates.

A key problem facing the field is that the ARGs estimated by heuristic large-scale 
methods lack a concrete connection to population genetic theory because
we cannot compute their likelihood under the SMC or similar models. 
Computing the likelihood of an ARG under a model is a fundamental element of
probabilistic inference. The classical Kuhner-Yamato-Felsenstein
\citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
works by assigning a probability to each recombination and common
ancestor event, and to the inter-event  waiting times. 
Although the KYF formulation is a natural and elegant way to describe 
the likelihood of an ARG, the high level of detail about the underlying
events that is required is not present 
in many estimated ARGs~\citep{wong_general_2023} or even in most simulations.
Simulators such as the classical ms
program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
which does not contain sufficient detail for likelihood computation.
The msprime
simulator~\citep{kelleher_efficient_2016,baumdicker_efficient_2021} 
by default outputs a more complete and efficient representation of 
the simulated ancestry, but still does not provide the exhaustive
detail required.
Indeed, providing the information required to 
support likelihood calculations originally motivated the addition of
the \texttt{record\_full\_arg}
option to msprime~\citep{baumdicker_efficient_2021}.
Forwards-time simulations that output 
ARGs~\citep{kelleher_efficient_2018,haller_tree_2018} also do not
typically contain the level of detail required to compute a likelihood under
the KYF approach.

@@

In this paper we provide a new formulation of ARG likelihood which 
has much more lenient input requirements than KYF, and gracefully handles
incomplete and imprecise ARGs.
% In this paper we provide a necessary first step towards this goal
% of statistically rigorous large-scale ARG inference, by defining a function
% to compute the likelihood of a general class of ARG under the SMC.
To do this, we give a backwards-in-time description of the SMC
which naturally leads to a straightforward method for computing the likelihood
of any ARG, and describe an efficient algorithm to do so.
We show that explicitly identifying recombination events via nodes 
in the graph is not necessary 
(from the perspective of computing the likelihood)
as long as the ``locally
unary''~\citep{wong_general_2023,fritze2024forest} sections of common ancestor
nodes are retained.
We also show via an empirical example that the likelihood is 
robust to the presence of polytomies and other imprecise 
information, and illustrate how it can be
decomposed into separately evaluated time slices.
Finally, we discuss how these results may be applied and extended.

@@

% % % % % % % % % % % % % % % % % %
\subsection*{ARGs} \label{par:recording}
Although the term Ancestral Recombination Graph (ARG) has been used to mean several things,
in the generic sense, it is a labelled graph structure that records
the inheritance of genetic material \citep{wong_general_2023}.
An ARG of the sort we consider describes the history of inheritance
of the genomic material of a focal set of individuals, or \emph{samples}.
Concretely, an ARG is equivalent to a collection of (non-contradictory) statements
of the form ``$c$ inherited from $p$ on the genomic segment from $x$ to $y$'',
where $c$ and $p$ are (haploid) genomes, and $x$ and $y$ are genetic coordinates.
We record each such statement in an ``edge'',
often summarised as the tuple $(p,c,x,y)$.
By ``describes the history'',
we mean that each node in the ARG represents a specific (ancestral) genome,
and that the ARG specifies the genealogical tree describing how the samples are related
at each point on the genome.

\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{figures/smc_custom_2rows_area_full_hap.png}
\caption{An ARG generated under the SMC, represented as a graph (top) and as a
series of local trees (bottom). To ensure the one-to-one correspondence
between both representations a node is recorded for each local tree whenever
a node is encountered along its ancestry path in the graph representation.
This results in nodes that (locally) have only a single child, and are therefore
(locally) unary. 
Nodes A, B and C are recombinants, i.e, have two parents in the graph
and are unary in the local trees. The blue shaded regions indicate
the range of possible times associated with these recombinants.
The blue arrows show the
standard left-to-right logic of the SMC whereby
the floating lineage coalesces randomly
with a the remaining portion of the local tree following recombination.
% By removing recombinant nodes, we limit the information on the time to each
% of these events to the windows (blue shaded array) delimited by the age of
% the child node and the minimum age of its parents.
}
\label{fig:smc-unary}
\end{figure}

An ARG can therefore be thought of either as a collection of inheritance
relationships between ancestral haplotypes,
or as a sequence of genealogical trees that have common node labels.
Fig.\ref{fig:smc-unary} shows a small ARG
that describes relationships between four samples on a 100bp piece of genome,
represented as both a graph (top) and a series of local trees (bottom).
Both structures represent the same information:
for instance, the fact that the ancestral genome at $A$
inherited the left half of its genome from ancestor $g$
and the right half from ancestor $f$
is seen in the fact that the first two trees have $A$ below $g$,
and the remaining have $A$ below $f$.
The blue dotted line from $A$ in the second tree to $f$ in the third tree
draws our attention to the fact that
the path traced by $A$'s ancestry switches at that point.
In fact, $A$ is in blue because it is a \emph{recombinant} node --
an ancestor that has a segment of genome inherited by a sample,
but that segment was inherited in two pieces from their parent.
Other nodes (labelled by black numbers) are either samples
or \emph{common ancestry} nodes --
ancestors that are the most recent common ancestor for two 
or more samples
at some place along the genome.

>>>>

% Responses to reviews
\input{review-response-commands}
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses
\reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
\newcommand{\responsefile}{review-responses}  % name of the review responses file

@@

\begin{abstract}
Ancestral recombination graphs (ARGs) are the focus of much ongoing research
interest. Recent progress in inference has made ARG-based approaches feasible
across of range of applications, and many new methods using inferred ARGs as
input have appeared. This progress on the long-standing problem of ARG
inference has proceeded in two distinct directions. First,
the Bayesian inference of ARGs under the Sequentially Markov
Coalescent (SMC), is now practical for tens-to-hundreds of samples. 
Second, approximate models and heuristics can now scale to sample sizes two to three
orders of magnitude larger. Although these heuristic methods are reasonably
accurate under many metrics, one significant drawback is that the ARGs they 
estimate do not have the topological properties required to compute a 
likelihood under models such as the SMC under present-day formulations.
In particular, heuristic inference methods typically do not estimate
precise details about recombination events, which are currently
required to compute a likelihood.
In this paper we present a 
backwards-time formulation of the SMC 
and derive a straightforward definition 
of the likelihood of a general class of ARG under this model. 
We show that this formulation does not require precise details of recombination events
to be estimated, and is robust to the presence of polytomies. 
We discuss the possibilities for inference that this opens. \revpoint{2}{3}
\end{abstract}

@@

\section*{Introduction}
Following recent breakthroughs in simulation and inference methods,
there is now strong interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary biology~\citep{lewanski_era_2024,brandt_promise_2024,nielsen_inference_2025}. \revpoint{2}{4}
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes. \revpoint{2}{5}
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Numerous applications taking advantage of this detailed history
in inferred ARGs are now appearing~\citep{
stern_approximate_2019,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
ignatieva_ongoing_2022,
wang_complex_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
osmond_estimating_2021,
huang_estimating_2024,
grundler_geographic_2024,
korfmann_simultaneous_2024,
deraje_inferring_2024,
speidel2025high}
and it seems likely that many more will
follow~\citep{harris_database_2019,harris_using_2023}.
Although there are many different approaches to ARG
inference~\citep{wong_general_2023}, two broad classes of method
have been the focus of most recent interest.

% TODO Maybe swap the order of these two paragraphs around, so it's 
% more natrual to talk about the advantages of model-driven stuff next.
The first class of inference methods are Bayesian approaches \revpoint{2}{6}
that sample ARGs under a population genetics model such as the
coalescent with recombination~\citep{hudson_properties_1983},
and its approximation, the Sequentially Markovian
Coalescent~\citep{mcvean_approximating_2005,marjoram_fast_2006}.
ARGweaver~\citep{rasmussen_genome-wide_2014,hubisz_mapping_2020} 
has been the most widely used and studied,
and until recently consistently out-performed other methods in terms 
of accuracy in a variety of benchmarks~\citep{brandt2022evaluation},
and has successfully been applied in several
contexts~\citep[e.g.,][]{de_chimpanzee_2016,shriner_whole_2018,hejase_genomic_2020,
stankowski_genetic_2024}. 
The recently-introduced method SINGER takes a similar approach,
and by using an improved Monte Carlo algorithm
promises to be even more accurate and substantially more 
scalable~\citep{deng_robust_2024}. 
The second class of inference methods that has been of recent interest 
are based on heuristics and approximate models.
Relate~\citep{speidel_method_2019},
tsinfer~\citep{kelleher_inferring_2019},
ARG-Needle~\citep{zhang_biobank-scale_2023} 
and Threads~\citep{gunnarsson_scalable_2024}
work on quite different 
principles, but share some common properties. 
Firstly, they are all 
heuristic and approximation driven, favouring computational
scalability over a basis in a statistical model. \revpoint{2}{7}
Secondly, they all produce a single
ARG as output, inferring a deterministic point estimate.
Thirdly, they regard the precise timing of nodes as a separate
problem, focusing only on the relative ordering of nodes when 
producing the initial ARG. General purpose ARG dating methods are 
appearing~\citep{wohns_unified_2022,deng2025general} and can also improve
dating performance in ARGs sampled from the SMC~\citep{deng2025general}.
Fourthly (and most importantly for the 
purposes of this paper) the ARGs that heuristic methods produce 
lack some of the topological information present in the output of ARGweaver or SINGER. 
This occurs, roughly speaking, because of a basic difference in their
approach to uncertainty.

There is considerable uncertainty in ARG inference, and often a fundamental
lack of information in the sequence data to distinguish different
possibilities. Consider, for example, a case in which we have three
ancestral lineages with identical sequences. We know that there
must have been two coalescence events, but there is no mutational information to 
distinguish their relative ordering. Sampling methods overcome this problem
by randomly choosing an ordering, with the uncertainty communicated by the order
changing in the different ARGs sampled from the posterior.
The other approach (used by methods such as tsinfer) is to communicate
the uncertainty structurally by means of a polytomy: we have no
information about the intermediate coalescent event, and so we omit it.
Similarly, there is often a fundamental lack of information about the 
details of recombination events, and heuristic methods (in different ways)
omit these details. \revpoint{2}{8}
In many ways, the significant progress made in
scalability by these heuristic methods is \emph{because} they don't
attempt to precisely reconstruct recombination events.

There are substantial advantages to the statistically rigorous sampling 
approaches of ARGweaver and SINGER.
In particular, they tend to be more accurate (at the 
scale at which they can be compared),
and sampling ARGs from a
posterior distribution provides a means of quantifying the uncertainty 
around estimates. \revpoint{2}{9}
On the other hand, heuristic methods have the potential
to use the greater amount of information in large datasets,
capturing fine-scale details about the recent past.
% thus possibly producing even more accurate estimates
% (although this has not, to our knowledge, been tested).
Although SINGER is a major step forward
in terms of scalability over ARGweaver and other methods, inference is still 
only feasible for hundreds of samples.  \revpoint{2}{1}
Meanwhile, tsinfer, ARG-Needle and Threads have 
been applied to datasets of hundreds of \emph{thousands} of samples.
With the explosive growth in dataset
sizes in recent years~\citep{caulfield2017national,turnbull2018100,
bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole,
all2024genomic,stark2024call,cook2025our}
there is a pressing need for more statistical
rigour in large scale ARG inference.
% This inability to connect the results of large-scale inference methods
% with a generative model
% is a fundamental limitation, and limits both the utility of 
% current point estimates as well as the possibilities for further
% inference. Specifically, we are unable to quantify uncertainty 
% around heuristic point estimates. We cannot explore ARG space
% around the point estimates using (e.g., MCMC), as we have no way definition
% of improvement. We cannot compare models of demography, or evaluate the 
% goodness of fit for model estimates.

A key problem facing the field is that the ARGs estimated by heuristic large-scale \revpoint{2}{10}
methods lack a concrete connection to population genetic theory because
we cannot compute their likelihood under the SMC or similar models. \revpoint{2}{11}
Computing the likelihood of an ARG under a model is a fundamental element of
probabilistic inference. The classical Kuhner-Yamato-Felsenstein
\citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
works by assigning a probability to each recombination and common
ancestor event, and to the inter-event  waiting times. 
Although the KYF formulation is a natural and elegant way to describe 
the likelihood of an ARG, the high level of detail about the underlying
events that is required is not present 
in many estimated ARGs~\citep{wong_general_2023} or even in most simulations.
Simulators such as the classical ms
program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
which does not contain sufficient detail for likelihood computation. \revpoint{2}{12}
The msprime
simulator~\citep{kelleher_efficient_2016,baumdicker_efficient_2021} 
by default outputs a more complete and efficient representation of 
the simulated ancestry, but still does not provide the exhaustive
detail required.
Indeed, providing the information required to 
support likelihood calculations originally motivated the addition of
the \texttt{record\_full\_arg}
option to msprime~\citep{baumdicker_efficient_2021}.
Forwards-time simulations that output 
ARGs~\citep{kelleher_efficient_2018,haller_tree_2018} also do not
typically contain the level of detail required to compute a likelihood under
the KYF approach.

@@

In this paper we provide a new formulation of ARG likelihood which 
has much more lenient input requirements than KYF, and gracefully handles
incomplete and imprecise ARGs. \revpoint{2}{13}
% In this paper we provide a necessary first step towards this goal
% of statistically rigorous large-scale ARG inference, by defining a function
% to compute the likelihood of a general class of ARG under the SMC.
To do this, we give a backwards-in-time description of the SMC
which naturally leads to a straightforward method for computing the likelihood
of any ARG, and describe an efficient algorithm to do so.
We show that explicitly identifying recombination events via nodes 
in the graph is not necessary 
(from the perspective of computing the likelihood)
as long as the ``locally
unary''~\citep{wong_general_2023,fritze2024forest} sections of common ancestor
nodes are retained.
We also show via an empirical example that the likelihood is 
robust to the presence of polytomies and other imprecise 
information, and illustrate how it can be
decomposed into separately evaluated time slices.
Finally, we discuss how these results may be applied and extended.

@@

% % % % % % % % % % % % % % % % % %
\subsection*{ARGs} \label{par:recording}
Although the term Ancestral Recombination Graph (ARG) has been used to mean several things,
in the generic sense, it is a labelled graph structure that records
the inheritance of genetic material \citep{wong_general_2023}.
An ARG of the sort we consider describes the history of inheritance
of the genomic material of a focal set of individuals, or \emph{samples}.
Concretely, an ARG is equivalent to a collection of (non-contradictory) statements
of the form ``$c$ inherited from $p$ on the genomic segment from $x$ to $y$'',
where $c$ and $p$ are (haploid) genomes, and $x$ and $y$ are genetic coordinates.
We record each such statement in an ``edge'',
often summarised as the tuple $(p,c,x,y)$.
By ``describes the history'',
we mean that each node in the ARG represents a specific (ancestral) genome, \revpoint{2}{14}
and that the ARG specifies the genealogical tree describing how the samples are related
at each point on the genome. \revpoint{2}{15}

\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{figures/smc_custom_2rows_area_full_hap.png}
\caption{An ARG generated under the SMC, represented as a graph (top) and as a
series of local trees (bottom). To ensure the one-to-one correspondence
between both representations a node is recorded for each local tree whenever
a node is encountered along its ancestry path in the graph representation.
This results in nodes that (locally) have only a single child, and are therefore
(locally) unary. 
Nodes A, B and C are recombinants, i.e, have two parents in the graph
and are unary in the local trees. The blue shaded regions indicate
the range of possible times associated with these recombinants.
The blue arrows show the
standard left-to-right logic of the SMC whereby
the floating lineage coalesces randomly
    with the remaining portion of the local tree following recombination.
% By removing recombinant nodes, we limit the information on the time to each
% of these events to the windows (blue shaded array) delimited by the age of
% the child node and the minimum age of its parents.
}
\label{fig:smc-unary}
\end{figure}

An ARG can therefore be thought of either as a collection of inheritance
relationships between ancestral haplotypes, \revpoint{2}{16}
or as a sequence of genealogical trees that have common node labels.
Fig.\ref{fig:smc-unary} shows a small ARG
that describes relationships between four samples on a 100bp piece of genome,
represented as both a graph (top) and a series of local trees (bottom).
Both structures represent the same information:
for instance, the fact that the ancestral genome at $A$
inherited the left half of its genome from ancestor $g$
and the right half from ancestor $f$
is seen in the fact that the first two trees have $A$ below $g$,
and the remaining have $A$ below $f$.
The blue dotted line from $A$ in the second tree to $f$ in the third tree
draws our attention to the fact that
the path traced by $A$'s ancestry switches at that point.
In fact, $A$ is in blue because it is a \emph{recombinant} node --
an ancestor that has a segment of genome inherited by a sample,
but that segment was inherited in two pieces from their parent.
Other nodes (labelled by black numbers) are either samples
or \emph{common ancestry} nodes -- \revpoint{2}{20}
ancestors that are the most recent common ancestor for two 
or more samples
at some place along the genome.

Crucial for the equivalence between the two representations is the presence of
nodes that only have a single child along one or more local trees --
i.e., are \emph{locally unary}~\citep{wong_general_2023}.
Recombinant nodes (in blue) have a single child in all trees.
However, common ancestry nodes can be locally unary as well:
for instance, sample $d$ inherited a longer stretch of genome from ancestor $f$
than sample $c$.
In the top plot, this is visible because the pink bar below node $f$
extends further to the left than does the blue bar,
while in the bottom plot, this is visible because node $f$ is parent to $d$
in the second tree, but not to node $c$, and so is unary. \revpoint{2}{21}
